---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# clptheory

<!-- badges: start -->
<!-- badges: end -->

The goal of clptheory (_classical price theory_) is to create a suite of functions to implement the classical theory of prices. The functions in this package computes the uniform rate of profit, the vector of price of production and the vector of labor values for different specifications of the circulating capital model and the capital stock model. The functions also computes various regression- and non-regression-based measures of deviation between the vector of _all possible_ relative prices of production and the vector of _all possible_ relative labor values.

## Installation

You can install the development version of clptheory from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("dbasu-umass/clptheory")
```

## Example of Circulating Capital Model

### Standard Interpretation
This is an example of a 3-industry economy which shows you how to: 
- compute the uniform rate of profit and the vectors of labor values and prices of production for a basic circulating capital model using the Standard Interpretation; and
- compute regression- and non-regression-based measures of deviation between the vector of _all possible_ relative prices of production and the vector of _all possible_ relative labor values.

```{r examplesi}
library(clptheory)

# ------ Data

# Input-output matrix
A <- matrix(
  data = c(0.265,0.968,0.00681,0.0121,0.391,0.0169,0.0408,0.808,0.165),
  nrow=3, ncol=3, byrow = TRUE
)

# Direct labor input vector
l <- matrix(
  data = c(0.193, 3.562, 0.616),
  nrow=1
)

# Real wage bundle vector
b <- matrix(
  data = c(0.0109, 0.0275, 0.296),
  ncol=1
)

# Gross output vector
Q <- matrix(
  data = c(26530, 18168, 73840),
  ncol=1
)

# Market prices vector
m <- matrix(
  data = c(4, 60, 7),
  nrow=1
)

# Uniform nominal wage rate
wavg <- m%*%b

# Vector of nominal wage rates
# (For simplicity, we use the uniform wage rate)
w <- matrix(
  data = rep(wavg,3),
  nrow = 1
)

# Estimate circulating capital model with SI
si1 <- ppstdint1(
  A = A,
  l = l,
  b = b,
  Q = Q,
  l_simple = l
)

```
What is the uniform rate of profit?
```{r}
si1$urop
```
What is the vector of labor values?
```{r}
si1$lvalues
```
What is the vector of prices of production (absolute)?
```{r}
si1$ppabs
```
Let us now compute the various non-regression-based measures of deviation between the vector of all possible relative labor values and the vector of all possible relative prices of production.
```{r}
nrsi1 <- nregtestrel(
  x = si1$ppabs,
  y = si1$lvalues,
  w = w,
  w_avg = wavg[1,1],
  mev = si1$mevg,
  Q = Q
)
(nrsi1)
```
Let us now compute the various non-regression-based measures of deviation between the vector of all possible relative labor values and the vector of all possible relative prices of production.
```{r}
rsi1 <- regtestrel(x=si1$ppabs,y=si1$lvalues)
(rsi1)
```


### New Interpretation
We continue with the 3-industry example to 
- compute the uniform rate of profit and the vectors of labor values and prices of production for a basic circulating capital model using the New Interpretation;
- and compute regression- and non-regression-based measures of deviation between the vector of _all possible_ relative prices of production and the vector of _all possible_ relative labor values.

```{r exampleni}
library(clptheory)

# ------ Data

# Input-output matrix
A <- matrix(
  data = c(0.265,0.968,0.00681,0.0121,0.391,0.0169,0.0408,0.808,0.165),
  nrow=3, ncol=3, byrow = TRUE
)

# Direct labor input vector
l <- matrix(
  data = c(0.193, 3.562, 0.616),
  nrow=1
)

# Real wage bundle vector
b <- matrix(
  data = c(0.0109, 0.0275, 0.296),
  ncol=1
)

# Gross output vector
Q <- matrix(
  data = c(26530, 18168, 73840),
  ncol=1
)

# Value of labor power
v <- 2/3

# Estimate circulating capital model with NI
ni1 <- ppnewint1(
  A = A,
  l = l,
  w = wavg[1,1],
  v = v,
  Q = Q,
  l_simple = l
)

```
What is the uniform rate of profit?
```{r}
ni1$urop
```
What is the vector of labor values?
```{r}
ni1$lvalues
```
What is the vector of prices of production (absolute)?
```{r}
ni1$ppabs
```
Let us now compute the various non-regression-based measures of deviation between the vector of all possible relative labor values and the vector of all possible relative prices of production.
```{r}
nrni1 <- nregtestrel(
  x=ni1$ppabs,
  y=ni1$lvalues,
  w=w,
  w_avg=wavg[1,1],
  mev=ni1$mevg,
  Q=Q
  )
(nrni1)
```
Let us now compute the various regression-based measures of deviation between the vector of all possible relative labor values and the vector of all possible relative prices of production.
```{r}
rni1 <- regtestrel(x=ni1$ppabs,y=ni1$lvalues)
(rni1)
```
We can compare the results from the SI and the NI for the non-regression-based measures of deviation between relative prices of production and relative values.
```{r}
comp1 <- cbind(nrsi1,nrni1)
colnames(comp1) <- c("SI","NI")
(comp1)
```

