---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# clptheory

<!-- badges: start -->
<!-- badges: end -->

The goal of `clptheory` (_classical price theory_) is to create a suite of functions to implement the classical theory of prices. The functions in this package computes the uniform rate of profit, the vector of price of production and the vector of labor values for different specifications of the circulating capital model and the capital stock model. The functions also computes various regression- and non-regression-based measures of deviation between the vector of _all possible_ relative prices of production and the vector of _all possible_ relative labor values.

## Installation

The easiest way to get `clptheory` is to install it from CRAN. Use the following in an R console:
```{r}
install.packages("clptheory")

```

You can install the development version of `clptheory` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("dbasu-umass/clptheory")
```

## Main Functions
This package provides the following functions.

1. `ppstdint1`: a function to estimate a basic circulating capital model (uniform wage rates across industries + not taking account of unproductive industries) with the Standard Interpretation;

2. `ppstdint2`: a function to estimate a circulating capital model (uniform wage rates across industries + takes account of unproductive industries) with the Standard Interpretation;

3. `ppstdint3`: a function to estimate a basic capital stock model (uniform wage rates across industries + not taking account of unproductive industries) with the Standard Interpretation;

4. `ppnewint1`: a function to estimate a basic circulating capital model (uniform wage rates across industries + not taking account of unproductive industries) with the New Interpretation;

5. `ppnewint2`: a function to estimate a circulating capital model (allows differential wage rates across industries + not taking account of unproductive industries) with the New Interpretation;

6. `ppnewint3`: a function to estimate a circulating capital model (uniform wage rates across industries + takes account of unproductive industries) with the New Interpretation;

7. `ppnewint4`: a function to estimate a circulating capital model (allows differential wage rates across industries + takes account of unproductive industries) with the New Interpretation;

8. `ppnewint5`: a function to estimate a basic capital stock model (uniform wage rates across industries + not taking account of unproductive industries) with the New Interpretation;

9. `ppnewint6`: a function to estimate a capital stock model (allows differential wage rates across industries + not taking account of unproductive industries) with the New Interpretation;

10. `ppnewint7`: a function to estimate a capital stock model (uniform wage rates across industries + takes account of unproductive industries) with the New Interpretation;

11. `ppnewint8`: a function to estimate a capital stock model (allows differential wage rates across industries + takes account of unproductive industries) with the New Interpretation;

12. `nregtestrel`: a function that computes various non-regression-based measures of deviation between the vector of _relative_ prices of production and the vector of _relative_ labor values;

13. `regtestrel`: a function that computes various regression-based measures of deviation between the vector of _relative_ prices of production and the vector of _relative_ labor values;

14. `createdata`: a function to create the data objects (matrices, vectors and scalars) necessary to implement the SI and NI.

The package contains the following three datasets.

1.  `aussea`: the socio economic accounts for the Australian economy extracted from the 2016 release of the World Input Output Database; this data set contains industry-level variables (53 industries) for the USA for 15 years, 2000-2014; 

2. `ausiot`: input-output tables for the Australian economy extracted from the 2016 release of the World Input Output Database; this data set contains 53-industry input-output tables for the USA for 15 years, 2000-2014;

3.  `usasea`: the socio economic accounts for the USA extracted from the 2016 release of the World Input Output Database; this data set contains industry-level variables (53 industries) for the USA for 15 years, 2000-2014; 

4. `usaiot`: input-output tables for the USA extracted from the 2016 release of the World Input Output Database; this data set contains 53-industry input-output tables for the USA for 15 years, 2000-2014;

5.  `usarwb`: personal consumption expenditure on the output of the 53 industries of the input-output tables for the USA extracted from the 2016 release of the World Input Output Database; this data set contains data for 15 years, 2000-2014. (Note: This data set is not necessary for the analysis.)

## Example 1: Analysis for Australia
Let us conduct price of production analysis for Australia (AUS) and see how to use the functions in `clptheory` to

1. compute the uniform rate of profit and the vectors of labor values and prices of production for a basic circulating capital model using the Standard Interpretation and the New Interpretation; and

2. compute regression- and non-regression-based measures of deviation between the vector of _all possible_ relative prices of production and the vector of _all possible_ relative labor values.

Let us load the package.
```{r}
# Load library
library(clptheory)
```


### Data
Let us create the data objects.
```{r}
ausdata <- clptheory::createdata(
  country = "AUS", year = 2010, 
  datasea = aussea, dataio = ausiot
  )
```

### Standard Interpretation
Let us now estimate the circulating capital model with SI.
```{r}
si1 <- clptheory::ppstdint1(
  A = ausdata$Ahat,
  l = ausdata$l,
  b = ausdata$b,
  Q = ausdata$Q,
  l_simple = ausdata$l_simple
)
```

### New Interpretation
Let us now estimate the circulating capital model with NI.
```{r}
ni1 <- clptheory::ppnewint1(
  A = ausdata$Ahat,
  l = ausdata$l,
  w = ausdata$wavg,
  v = ausdata$vlp,
  Q = ausdata$Q,
  l_simple = ausdata$l_simple
)
```
Let us see the uniform profit rate.
```{r}
cbind(si1$urop,ni1$urop)
```

### Non-Regression-Based Measures of Deviation
Let us compute various non-regression-based measures of the deviation
between the vector of *relative* labor values and the vector of
*relative* prices of production for the SI.

```{r}
nrsi1 <- clptheory::nregtestrel(
  x = si1$ppabs,
  y = si1$lvalues,
  w = ausdata$wagevector_all,
  w_avg = ausdata$wavg,
  mev = si1$mevg,
  Q = ausdata$Q
)
```

Let us do the same computation for the NI.

```{r}
nrni1 <- clptheory::nregtestrel(
  x=ni1$ppabs,
  y=ni1$lvalues,
  w=ausdata$wagevector_all,
  w_avg=ausdata$wavg,
  mev=ni1$mevg,
  Q=ausdata$Q
  )
```

We can now compare the results for the analysis of the _circulating
capital model_ from the SI approach and the NI approach for the
non-regression-based measures of deviation between relative prices of
production and relative values.

```{r}
comp1 <- cbind(nrsi1,nrni1)
colnames(comp1) <- c("SI","NI")
(comp1)
```

In the results above, we see the magnitudes of six different measures of
the deviation between the vector of *relative* prices of production and
the vector of *relative* labor values: root mean squared error (RMSE),
meann absolute distance (MAD), mean absolute weighted distance (MAWD),
classical distance measure (CDM), angle between the two vectors (angle
in degrees), and distance computed using angle (distance).

As an example, we can see that the CDM for SI is 0.789 and for NI is
0.647. This can be interpreted as showing that the deviation between the
vector of *relative* prices of production and the vector of *relative*
labor values is 79 percent and 65 percent of the relative value vector
according to the SI and NI methodology, respectively.

The last row of the above results shows the length of (number of
observations in) the vector of *relative* prices of production or the
vector of *relative* labor values. Recall that the input-output matrix
is 48 by 48. Hence, the *absolute* value and price of production vectors
will each be of size 48. Thus, the size of the vector of *relative*
prices of production (or labor value) should be $(48 \times 47)/2=1128$.

### Regression-Based Measures of Deviation
Let us compute various regression-based measures of the deviation
between the vector of *relative* labor values and the vector of
*relative* prices of production for the SI.

```{r}
rsi1 <- clptheory::regtestrel(
  x = si1$ppabs,
  y = si1$lvalues
)
```

Let us do the same computation for the NI.

```{r}
rni1 <- clptheory::regtestrel(
  x=ni1$ppabs,
  y=ni1$lvalues
  )
```

We can now compare the results for the analysis of the *circulating
capital model* from the SI approach and the NI approach for the
regression-based measures of deviation between relative prices of
production and relative values.

```{r}
comp2 <- cbind(rsi1,rni1)
colnames(comp2) <- c("SI","NI")
(comp2)
```
Regression-based tests of the deviation between use regressions, either log-log or level-level, of relative prices of production on relative labor value. The key the null (joint) hypothesis is that the intercept is 0 and the slope is 1.

The F-stat in the log-log regression of relative prices of production on relative value is 143.91 for SI and 550.97 for NI. In both cases, we can strongly reject the null hypothesis that the intercept is 0 and the slope is 1. The corresponding F-stats for the level-level regressions are 35.75 (SI) and 266.39 (NI). Once again, the null hypothesis is strongly rejected.


## Example 2: Simple 3-Industry Set-up
This example was presented on pages 46-57 of E. M. Ochoa's dissertation (Ochoa, E. M. 1984. Labor-Value and Prices of Production: An Interindustry Study of the U.S. Economy, 1947–1972. PhD thesis, _New School for Social Research_, New York, NY.). This example has also been discussed in Appendix B of Basu and Moraitis, 2023. (Basu, Deepankar and Moraitis, Athanasios, "Alternative Approaches to Labor Values and Prices of Production: Theory and Evidence" (2023). Economics Department Working Paper Series. 347. _UMass Amherst_. URL: https://scholarworks.umass.edu/econ_workingpaper/347/)

### The Data
Let us load the library and create the data for our examples.
```{r data}
# Input-output matrix
A <- matrix(
  data = c(0.265,0.968,0.00681,0.0121,0.391,0.0169,0.0408,0.808,0.165),
  nrow=3, ncol=3, byrow = TRUE
)

# Depreciation matrix
D <- matrix(
  data = c(0,0,0,0.00568,0.0267,0.0028,0.00265,0.0147,0.00246),
  nrow=3, ncol=3, byrow = TRUE
)

# Direct labor input vector
l <- matrix(
  data = c(0.193, 3.562, 0.616),
  nrow=1
)

# Real wage bundle vector
b <- matrix(
  data = c(0.0109, 0.0275, 0.296),
  ncol=1
)

# Gross output vector
Q <- matrix(
  data = c(26530, 18168, 73840),
  ncol=1
)

# Market prices vector
m <- matrix(
  data = c(4, 60, 7),
  nrow=1
)

# Capital stock coefficient matrix
K <- matrix(
  data = c(0,0,0,0.120,0.791,0.096,0.037,0.251,0.043),
  nrow=3, ncol=3, byrow = TRUE
)

# Diagonal matrix of turnover times
t <- diag(c(0.317, 0.099, 0.187))

# Uniform nominal wage rate
wavg <- m%*%b

# Vector of nominal wage rates
w <- matrix(
  data = rep(wavg,3),
  nrow = 1
)

# Value of labor power
v <- 2/3

```


### Standard Interpretation
We will first implement the classical theory of prices for the circulating capital model and then turn to the capital stock model.

#### Circulating Capital Model
```{r examplesi}

# Estimate circulating capital model with SI
si1 <- ppstdint1(
  A = A,
  l = l,
  b = b,
  Q = Q,
  l_simple = l
)

```
What is the uniform rate of profit?
```{r}
si1$urop
```
What is the vector of labor values?
```{r}
si1$lvalues
```
What is the vector of prices of production (absolute)?
```{r}
si1$ppabs
```
Let us now compute the various non-regression-based measures of deviation between the vector of all possible relative labor values and the vector of all possible relative prices of production.
```{r}
nrsi1 <- nregtestrel(
  x = si1$ppabs,
  y = si1$lvalues,
  w = w,
  w_avg = wavg[1,1],
  mev = si1$mevg,
  Q = Q
)
```

#### Capital Stock Model
```{r}
# Estimate model
si2 <- ppstdint3(
  A = A,
  l = l,
  b = b,
  Q = Q,
  D = D,
  K = K,
  t = t,
  l_simple = l
)
```
What is the uniform rate of profti?
```{r}
si2$urop
```
What is the vector of labor values?
```{r}
si2$lvalues
```
What is the vector of prices of production?
```{r}
si2$ppabs
```
Let us now compute the non-regression-based measures of deviation.
```{r}
nrsi2 <- nregtestrel(
  x = si2$ppabs,
  y = si2$lvalues,
  w = w,
  w_avg = wavg[1,1],
  mev = si2$mevg,
  Q = Q
)
```


### New Interpretation
We continue working with the 3-industry example and implement the New Interpretation of Marx's labor theory of value. 

#### Circulating Capital Model
```{r exampleni}

# Estimate circulating capital model with NI
ni1 <- ppnewint1(
  A = A,
  l = l,
  w = wavg[1,1],
  v = v,
  Q = Q,
  l_simple = l
)

```
What is the uniform rate of profit?
```{r}
ni1$urop
```
What is the vector of labor values?
```{r}
ni1$lvalues
```
What is the vector of prices of production (absolute)?
```{r}
ni1$ppabs
```
Let us now compute the various non-regression-based measures of deviation between the vector of all possible relative labor values and the vector of all possible relative prices of production.
```{r}
nrni1 <- nregtestrel(
  x=ni1$ppabs,
  y=ni1$lvalues,
  w=w,
  w_avg=wavg[1,1],
  mev=ni1$mevg,
  Q=Q
  )
```

#### Capital Stock Model
```{r}
ni2 <- ppnewint5(
  A = A,
  l = l,
  v = v,
  w = wavg[1,1],
  Q = Q,
  D = D,
  K = K,
  t = t,
  l_simple = l
)
```
What is the uniform rate of profit?
```{r}
ni2$urop
```
What is the vector of labor values?
```{r}
ni2$lvalues
```
What is the vector of prices of production (absolute)?
```{r}
ni2$ppabs
```
Let us now compute the various non-regression-based measures of deviation between the vector of all possible relative labor values and the vector of all possible relative prices of production.
```{r}
nrni2 <- nregtestrel(
  x=ni2$ppabs,
  y=ni2$lvalues,
  w=w,
  w_avg=wavg[1,1],
  mev=ni2$mevg,
  Q=Q
  )

```

## Comparison of SI and NI
We can compare the results for the analysis of the _circulating capital model_ from the SI approach and the NI approach for the non-regression-based measures of deviation between relative prices of production and relative values.
```{r}
comp1 <- cbind(nrsi1,nrni1)
colnames(comp1) <- c("SI","NI")
(comp1)
```

We can compare the results for the analysis of the _capital stock model_ from the SI approach and the NI approach for the non-regression-based measures of deviation between relative prices of production and relative values.
```{r}
comp2 <- cbind(nrsi2,nrni2)
colnames(comp2) <- c("SI","NI")
(comp2)
```

## References

-   Basu, D. and Moraitis, T. (2023). Alternative Approaches to Labor
    Values and Prices of Production: Theory and Evidence. *Economics
    Department Working Paper Series*. 347. University of Massachusetts
    Amherst. Retrieved from
    <https://scholarworks.umass.edu/econ_workingpaper/347/>
