---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# clptheory

<!-- badges: start -->
<!-- badges: end -->

The goal of clptheory (_classical price theory_) is to create a suite of functions to compute the uniform rate of profit, the vector of price of production and the vector of labor values for different specifications of the circulating capital model and the capital stock model.

## Installation

You can install the development version of clptheory from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("dbasu-umass/clptheory")
```

## Example

### Standard Interpretation
This is a basic example which shows you how to compute the uniform rate of profit and the vectors of labor values and prices of production for a basic circulating capital model using the Standard Interpretation:

```{r examplesi}
library(clptheory)

# ------ Data

# Input-output matrix
A <- matrix(
  data = c(0.265,0.968,0.00681,0.0121,0.391,0.0169,0.0408,0.808,0.165),
  nrow=3, ncol=3, byrow = TRUE
)

# Direct labor input vector
l <- matrix(
  data = c(0.193, 3.562, 0.616),
  nrow=1
)

# Real wage bundle vector
b <- matrix(
  data = c(0.0109, 0.0275, 0.296),
  ncol=1
)

# Gross output vector
Q <- matrix(
  data = c(26530, 18168, 73840),
  ncol=1
)

# Estimate circulating capital model with SI
si1 <- ppstdint1(
  A = A,
  l = l,
  b = b,
  Q = Q,
  l_simple = l
)

```
What is the uniform rate of profit?
```{r}
si1$urop
```
What is the vector of labor values?
```{r}
si1$lvalues
```
What is the vector of prices of production (absolute)?
```{r}
si1$ppabs
```

### New Interpretation
This is a basic example which shows you how to compute the uniform rate of profit and the vectors of labor values and prices of production for a basic circulating capital model using the New Interpretation:

```{r exampleni}
library(clptheory)

# ------ Data

# Input-output matrix
A <- matrix(
  data = c(0.265,0.968,0.00681,0.0121,0.391,0.0169,0.0408,0.808,0.165),
  nrow=3, ncol=3, byrow = TRUE
)

# Direct labor input vector
l <- matrix(
  data = c(0.193, 3.562, 0.616),
  nrow=1
)

# Real wage bundle vector
b <- matrix(
  data = c(0.0109, 0.0275, 0.296),
  ncol=1
)

# Gross output vector
Q <- matrix(
  data = c(26530, 18168, 73840),
  ncol=1
)

# Market prices vector
m <- matrix(
  data = c(4, 60, 7),
  nrow=1
)

# Uniform nominal wage rate
wavg <- m%*%b

# Value of labor power
v <- 2/3

# Estimate circulating capital model with NI
ni1 <- ppnewint1(
  A = A,
  l = l,
  w = wavg[1,1],
  v = v,
  Q = Q,
  l_simple = l
)

```
What is the uniform rate of profit?
```{r}
ni1$urop
```
What is the vector of labor values?
```{r}
ni1$lvalues
```
What is the vector of prices of production (absolute)?
```{r}
ni1$ppabs
```


