---
title: "Prices of Production and Labor Values"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prices of Production and Labor Values}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Circulating Capital Model

<!-- badges: start -->

<!-- badges: end -->

Consider a circulating capital model of a capitalist economy, where
technology of production is captured by the $n \times n$ input-output
matrix, $\hat{A}$, and the $1 \times n$ direct labor input vector,
$\hat{l}$. The $i,j$-th element of the input-output matrix,
$\hat{a}_{ij}$, and the $i$-th element of the direct labor vector,
$\hat{l}_i$, denote, respectively, the dollars of commodity $i$ and the
amount of direct labor input (e.g., measured in hours) needed to produce
one dollar of commodity $j$.

The direct labor input vector, $\hat{l}$, is the vector of what the
Marxian tradition calls *complex labor*. This labor input registers the
differentials in average levels of skills and training associated with
the labor inputs across industries. Thus, the direct labor input vector,
measured in hours, will represent hours of labor of different degrees of
complexity. We can contrast complex labor with *simple labor*, which is
the labor performed by workers with basic skills available at any point
in time to all members of a society. Complex labor, then, requires
additional skill acquisition and training, in comparison to simple
labor. We can convert the vector of complex labor, $\hat{l}$, into the
vector of simple labor, $\hat{l}_s$, by multiplying each element of the
former with the ratio of that industry's average wage rate and the
minimum of the average nominal wage rate across industries, i.e.
\begin{equation}
        \hat{l}_{s,i} = \left( \frac{w_i}{w_{min}}\right)  \hat{l}_i,
    \end{equation} where $i=1,2, \ldots, n$, denotes the $i$-th industry
and the minimum is taken over all $i$.

### Labor Value

The vector of labor values of the commodities is given by
\begin{equation}\label{value-def1}
        \lambda = \lambda \hat{A} + \hat{l}_s,
    \end{equation} where $\lambda$ is the labor value vector per US
dollar (or any other national currency) of gross output.

### Price of Production

Mobility of capital across the $n$ industries will give rise to a long
run tendency for the rate of profit to equalize across industries. The
set of long run equilibrium prices that ensure the same, uniform rate of
profit across industries are known as prices of production. The
$1 \times n$ vector of prices of production, $p$, is given by
\begin{equation}
        p = \left( 1+r\right) \left( p \hat{A} + w \hat{l}\right), 
    \end{equation} where $p$ is the price of production vector per US
dollar (or any other national currency) of gross output.

### Standard Interpretation

Scholars following the standard interpretation (SI) take the
$n \times 1$ vector of *nominal* consumption expenditure on the $n$
commodities by the average worker per hour of labor input, $\hat{b}$,
*as given*. Using the classical savings assumption, they posit that
workers spend their wage income completely on consumption, so that
\begin{equation}
        w = p \hat{b}, 
    \end{equation} where $B$ is the $n \times 1$ vector of total
(personal) consumption and $L = \sum_i L_i$ is the (scalar) sum of total
labor inputs (measured in hours) across all industries. Substituting the
expression for $w$ in the price of production system, we get
\begin{equation}
        p = \left( 1+r\right) p \hat{M},  
    \end{equation} where \begin{equation}\label{def-M}
        \hat{M} = \hat{A} + \hat{b} \hat{l}
    \end{equation} is the augmented input matrix per US dollar of gross
output.

If $\hat{M}$ is nonnegative and irreducible, then we can apply the
Perron-Frobenius theorem to get \begin{equation}
        r = \frac{1}{\rho (\hat{M}) } - 1,
    \end{equation} where $\rho (\hat{M})$ refers to the maximal
eigenvalue of the matrix $\hat{M}$. The eigenvector of $\hat{M}$
corresponding to $\rho (\hat{M})$ gives the *relative* price of
production vector, $p_{rel}$. The vector of *absolute* prices of
production is then arrived at by multiplying the relative price vector
with the following constant: \begin{equation}
    \kappa = \frac{X}{p_{rel}X}.
    \end{equation} Here, $\kappa$ is the ratio of the nominal gross
output vector, $X$, aggregated with the vector market prices (the
numerator) and the same vector, $X$, aggregated with the relative price
vector (the denominator). In the numerator, the vector of market prices
is just the unit vector--because $X$ is already expressed in market
prices. Thus, \begin{equation}
        p^{SI} = \kappa p_{rel},
    \end{equation} where $p^{SI}$ is the $1 \times n$ vector of absolute
prices of production estimated by following the SI.

### New Interpretation

In the New Interpretation (NI) of Marx's labor theory of value, the
value of labor power, $v$, and the uniform nominal wage rate, $w$, are
*taken as given*, instead of the consumption bundle $\hat{b}$. The value
of labor power is the product of the nominal wage rate and the value of
money: \begin{equation}
        v = w \times \mu^{NI}.
    \end{equation} In contrast to the SI, the NI defines the value of
money with the net output vector measured in monetary units, $Y$:
\begin{equation}
        \mu^{NI} = \frac{\lambda Y}{p^{NI} Y} =  \frac{\hat{l}_s \left( I-\hat{A}\right)^{-1}  Y}{p^{NI} Y} = \frac{\hat{l}_s X}{p^{NI} Y},
    \end{equation} where $Y$ and $X$ are the $n \times 1$ vectors of net
and gross outputs in monetary units, $p^{NI}$ and $\mu^{NI}$ denote,
respectively, the absolute price of production vector and the value of
money, according to the NI. This definition embeds an invariance
principle with respect to the net output: the vector of net output
aggregated with the prices of production and multiplied with the value
of money (to covert it into units of labor hours) gives the same
magnitude as when the net output vector is aggregated with labor values.
The price of production system can, in turn, be written as
\begin{equation}
        p^{NI} = \left( 1+r\right) \left( p^{NI} \hat{A} + w \hat{l}\right), 
    \end{equation} and so the NI can be captured by the above two
equation systems. We can now solve this equation system for the uniform
rate of profit, $r$, and the price of production vector, $p^{NI}$, using
the 3-step algorithm given in Appendix 7.A.4 in Basu (2021).

The first step of the algorithm involves computing the maximal rate of
profit, $R$. If $\hat{A}$ is nonnegative and irreducible, then we can
use the Perron-Frobenius theorem on the price of production system of
equations to get $R = (1/\rho(\hat{A}))-1$, where $\rho(\hat{A})$ is the
maximal eigenvalue of the matrix $A$.

In the second step of the algorithm, we define the following univariate
function of $r$, \begin{equation}
        f(r) = \left( 1+r\right)w\hat{l}\left[ I - \left( 1+r\right) \hat{A}\right]^{-1}\left( I - \hat{A}\right)X - \frac{w \hat{l}_s X}{v},   
    \end{equation} and find the unique zero of this function over the
range, $0 \leq r < R$, i.e. the value of $r=r^*$ for which $f(r^*)=0$;
this gives us the uniform rate of profit, $r^*$. To understand why, note
that the price of production system gives us $$
    p^{NI} = \left( 1+r\right)w\hat{l}\left[ I - \left( 1+r\right) \hat{A}\right]^{-1}
    $$ and since, using the definition of the value of labor power, we
have $\mu^{NI} w = v$, using the expression for $\mu^{NI}$ and the
expression for the price of production above, we get $$
    \left( 1+r\right)w\hat{l}\left[ I - \left( 1+r\right) \hat{A}\right]^{-1}\left( I - \hat{A}\right)X = \frac{w \hat{l}_s X}{v}.
    $$ This implies that the uniform rate of profit is the zero of the
univariate function $f(.)$. To see why this is unique, note that if the
value of labor power, $v$, satisfies, $$
        0< v< \frac{\hat{l}_sX}{\hat{l}X},
        $$ where $X$ is the vector of nominal gross outputs, then
$f(0)=w (\hat{l} X- (\hat{l}_s X/v))<0$, $\lim_{r \to R} f(r) = +\infty$
and $f'>0$. That is why there is a unique value of $r=r^*>0$ at which
$f(r^*)=0$.

In the third, and final, step of the algorithm, we solve for the
absolute price of production vector as $$
    p^{NI} = \left( 1+r^*\right)w\hat{l}\left[ I - \left( 1+r^*\right) \hat{A}\right]^{-1}
    $$ using the uniform rate of profit that we got in the previous
step.

## Implementing SI and NI

Let us now use the functions and data sets available in the package
`clptheory` to implement the SI and NI methodologies to computing prices
of production, uniform rate of profit and labor values.

## The Data Set

Let us load the package

```{r}
library(clptheory)

```

and note that data sets for Australia (AUS) and the United States of
America (USA) are included in the package:

-   `aussea`: the socio economic accounts for AUS of the 2016 release of
    the World Input Output Database (WIOD);

-   `ausiot`: the input-output table for AUS of the 2016 release of the
    World Input Output Database (WIOD);

-   `usasea`: the socio economic accounts for USA of the 2016 release of
    the World Input Output Database (WIOD);

-   `usaiot`: the input-output table for USA of the 2016 release of the
    World Input Output Database (WIOD).

There is an additional data set that we will not use:

-   `usarwb`: the personal consumption expenditure vector for the USA of
    the 2016 release of the World Input Output Database (WIOD).

The socio economic accounts provide industry-level data on many
macroeconomic variables like employment, value added, gross output,
employee compensation. All variables, other than employment, are in
units of millions of US dollars; employment and hours are in units of
thousands. Let us see the names of all the variables in the socio
economic accounts data set.

```{r}
names(aussea)
```

The input-output tables provide data on the transaction flows between
industries, i.e. how much inputs does any industry purchase from any
other industry. Entries in the input-output table are in units of
millions of local currency. As an example, let us extract the input
output table for Australia for 2010 and see its dimensions.

```{r}
dim(ausiot[ausiot$Year==2010,])
```

It is not yet a square matrix because the columns on the far left are
for year, industry code, etc; and the columns in the far right are the
aggregate expenditure categories like personal consumption, investment,
etc. Using the `createdata` function below, we will be able to extract
the relevant columns and rows to ensure that it becomes a square matrix.

## Analysis for Australia (AUS)

Let us create the data objects for AUS for the year 2010 (we can choose
any year between 2000 and 2014).

```{r}
ausdata <- clptheory::createdata(
  country = "AUS", year = 2010, 
  datasea = aussea, dataio = ausiot
  )
```

The function `createdata` creates `ausdata` as a list

```{r}
class(ausdata)

```

which contains all the data objects that we need.

```{r}
names(ausdata)
```

In this object we have: `Ahat`, the input-output matrix; `l`, the direct
(complex) labor input vector; `l_simple`, the direct (simple labor input
vector); `Q`, the gross output vector; `wavg`, the uniform (average)
nominal wage rate; `wagevector_all`, the vector of nominal wage rates;
`vlp`, the value of labor power; and `pshare`, the average profit share.

What is the dimension of the the input-output matrix?

```{r}
dim(ausdata$Ahat)
```

It is indeed a *square* matrix and is of size 48 by 48.

### Standard Interpretation

Since the WIOD does not contain data on capital stock, we can only
estimate the circulating capital model. Let us estimate the circulating
capital model with SI.

```{r}
si1 <- clptheory::ppstdint1(
  A = ausdata$Ahat,
  l = ausdata$l,
  b = ausdata$b,
  Q = ausdata$Q,
  l_simple = ausdata$l_simple
)
```

The function `ppstdint1` implements the SI approach for a basic
circulating capital model and the output is a list.

```{r}
class(si1)
names(si1)
```

In the object `si1` we have: `meig`, the maximum eigenvalue of the
augmented input-output matrix, $M$; `urop`, the uniform rate of profit;
`mrop`, the maximal rate of profit (if the wage rate were zero);
`ppabs`, the vector of absolute prices of production; `pprel`, the
vector of relative prices of production (which is `ppabs` divided by the
first element of `ppabs`); `lvalues`, the vector of labor values;
`dprice`, the vector of direct prices, i.e. prices proportional to labor
values; and `mevg`, the monetary expression of value using the gross
output; `mnonneg`, dummy variable for whether M is nonnegative (1 =
nonnegative; 0 = not nonnegative); `mirred`, dummy variable for whether
M is irreducible (1 = irreducile; 0 = not irreducible).

### New Interpretation

Let us now estimate the circulating capital model with NI.

```{r}
ni1 <- clptheory::ppnewint1(
  A = ausdata$Ahat,
  l = ausdata$l,
  w = ausdata$wavg,
  v = ausdata$vlp,
  Q = ausdata$Q,
  l_simple = ausdata$l_simple
)
```

The function `ppnewint1` implements the NI approach for a basic
circulating capital model and the output is a list.

```{r}
class(ni1)
names(ni1)
```

In the object `ni1` we have: `meig`, the maximum eigenvalue of the
input-output matrix, $A$; `urop`, the uniform rate of profit; `mrop`,
the maximal rate of profit (if the wage rate were zero); `ppabs`, the
vector of absolute prices of production; `pprel`, the vector of relative
prices of production (which is `ppabs` divided by the first element of
`ppabs`); `lvalues`, the vector of labor values; `mevn`, the monetary
expression of value using the net output; `mevg`, the monetary
expression of value using the gross output; `Anonneg`, dummy variable
for whether A is nonnegative (1 = nonnegative; 0 = not nonnegative); and
`Airred`, dummy variable for whether A is irreducible (1 = irreducile; 0
= not irreducible).

### Results: Uniform Rate of Profit

Let us see the uniform profit rate.

```{r}
cbind(si1$urop,ni1$urop)
```

### Results: Deviation between Labor Values and Prices of Production

Let us compute various non-regression-based measures of the deviation
between the vector of *relative* labor values and the vector of
*relative* prices of production for the SI.

```{r}
nrsi1 <- clptheory::nregtestrel(
  x = si1$ppabs,
  y = si1$lvalues,
  w = ausdata$wagevector_all,
  w_avg = ausdata$wavg,
  mev = si1$mevg,
  Q = ausdata$Q
)
```

Let us do the same computation for the NI.

```{r}
nrni1 <- clptheory::nregtestrel(
  x=ni1$ppabs,
  y=ni1$lvalues,
  w=ausdata$wagevector_all,
  w_avg=ausdata$wavg,
  mev=ni1$mevg,
  Q=ausdata$Q
  )
```

We can now compare the results for the analysis of the *circulating
capital model* from the SI approach and the NI approach for the
non-regression-based measures of deviation between relative prices of
production and relative values.

```{r}
comp1 <- cbind(nrsi1,nrni1)
colnames(comp1) <- c("SI","NI")
(comp1)
```

In the results above, we see the magnitudes of six different measures of
the deviation between the vector of *relative* prices of production and
the vector of *relative* labor values: root mean squared error (RMSE),
meann absolute distance (MAD), mean absolute weighted distance (MAWD),
classical distance measure (CDM), angle between the two vectors (angle
in degrees), and distance computed using angle (distance).

As an example, we can see that the CDM for SI is 0.789 and for NI is
0.647. This can be interpreted as showing that the deviation between the
vector of *relative* prices of production and the vector of *relative*
labor values is 79 percent and 65 percent of the relative value vector
according to the SI and NI methodology, respectively.

The last row of the above results shows the length of (number of
observations in) the vector of *relative* prices of production or the
vector of *relative* labor values. Recall that the input-output matrix
is 48 by 48. Hence, the *absolute* value and price of production vectors
will each be of size 48. Thus, the size of the vector of *relative*
prices of production (or labor value) should be $(48 \times 47)/2=1128$.

## Analysis for USA

Let us create the data objects for USA for the year 2010 (we can choose
any year between 2000 and 2014).

```{r}
usadata <- clptheory::createdata(
  country = "USA", year = 2010, 
  datasea = usasea, dataio = usaiot
  )
```

Let us test if $M$ is irreducible.

```{r}
M <- (usadata$Ahat + usadata$b%*%usadata$l)
popdemo::isIrreducible(M)
```

Hence, we will not be able to implement the SI for the 2010 data set for
USA.

## References

-   Basu, D. and Moraitis, T. (2023). Alternative Approaches to Labor
    Values and Prices of Production: Theory and Evidence. *Economics
    Department Working Paper Series*. 347. University of Massachusetts
    Amherst. Retrieved from
    <https://scholarworks.umass.edu/econ_workingpaper/347>
